---
title: "Intro to Data Analysis in R - Sorting and Filtering"
author: "Aaron Kessler"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load the packages we'll need
# Remember, think of this like packing for our trip
# (We don't take the whole closet, only what we need)
library(tidyverse)
library(writexl)
library(nycflights13)

```

## Using tables in R

In R, data tables are known as dataframes.

What are "tibbles"...?

They're dataframes, with some additional tidyverse-infused features. Returns more readable output in the console, etc.

## Looking at some data - NY flights

We'll begin with a sample dataset on flights in and out of NYC's three airports in 2013 - included in the package we loaded above.

```{r}

flightdata <- nycflights13::flights

flightdata

```

What's with that arrow? Let's explain.

It's how you "assign" a stored value in R. How you save something with a name of your choosing, instead of just spit it out to the current window itself in the R session.

You saw above that if we want to see a dataframe/tibble, we can feed R just its name and it gets returned - the whole thing, so to speak.

We can also return just the few rows using `head()`.

```{r}

head(flightdata)

```

We can also see a vertical list of all the columns using `glimpse()`.

```{r}

glimpse(flightdata)

```

## Filtering and Sorting

There's a lot of flight here. So let's talk about how we can use tidyverse functions to filter and sort our data, just like you may be used to in other tools like Excel, SQL or other methods.

Let's say I only want to see flights that were going to Sarasota, Florida (SRQ).

```{r}
 
filter(flightdata, dest == "SRQ")

```

Why the double-equal-sign? Let's talk about it.

For numeric values, you can also use boolean operators.

```{r}

filter(flightdata, month <= 3)

```

Wow, that's still a lot of stuff - that's because we're showing *all* flights in the entire table that were in the first quarter of 2013.

But wait, didn't we just tell R before that we wanted *only the ones from Sarasota*? What gives.

That's because in R, you're never actually changing the original table, unless you explicitly "save" it as a new object. So since our filter above only told R to filter by month, that's what it did.

Now, let's actually filter by more than one thing at a time, so we can capture both our restrictions at once.

```{r}

filter(flightdata, dest == "SRQ", month <= 3)
#note the comma separating the filtering terms

```

There we go.

But there's an even better way to write this out...

### Why I stopped worrying and learned to love the "pipe"

This can still get a little confusing once you wind up with larger amounts of steps to string together.

Enter a glorious feature of the tidyverse: the PIPE. It looks like this `%>%`

The "pipe" (shortcut is CTRL/CMD + SHIFT + M) allows you to chain together commands, in a way that can be more intuitive for how humans thing through each step we're trying to do.

```{r}

flightdata %>% 
  filter(dest == "SRQ")

```

Voila! So what just happened there?

Think of `%>%` as the equivalent of *"and then do this"*...

It takes the result and applies something new to it, in sequential order. This becomes easier to actually see when we add new functions to the chain- so let's talk about *sorting* our data with `arrange()`

```{r}

flightdata %>% 
  arrange(dep_delay)

```

To sort by highest value, add `desc()` - for descending order.

```{r}

flightdata %>% 
  arrange(desc(dep_delay))

```

Ok, so now that we know how to sort things, let's use the pipe to chain together all of the steps...and see what it looks like:

```{r}

flightdata %>% 
  filter(dest == "SRQ", month <= 3) %>% 
  arrange(desc(dep_delay))

```

Bingo, now we have the flights only to Sarasota, in the first quarter of the year, ordered by which flights had the longest delays in their departures.

Think about what we just did here. You can read the code out loud, and it intuitively makes sense.

To see the difference, this is how you would have written the same request to R without the pipe:

```{r}

arrange(filter(flightdata, dest == "SRQ", month <= 3), desc(dep_delay))

```

It's almost like you have to think backwards. But this is actually how many programming languages work, where you have to structure (and read) things almost inside-out. And the example above is still relatively simple...imagine what starts happening when you add all kinds of other steps to the mix.

It can become very difficult to think through the steps when you're writing what you want when you have to think inside-out rather than sequentially. What's more, it can be very difficult for other people to read your code - not to mention harder for you to read your own code later on..!

## Limiting to only certain columns

One more thing - what if we don't want all the columns? Just some of them. That flight dataset has a whole bunch of stuff, and maybe we don't need it all depending on our particular goal for an analysis.

This happens all the time.

It's also frequently the case where you'd like to change the order of columns of your choosing.

Dplyr makes this all easy using the `select()` function. Let's take a look...

```{r}

flightdata  %>% 
  select(carrier, flight, origin, dest, month, day, sched_dep_time, dep_delay)

```

You can also do the reverse, where you tell it which columns you'd *don't* want.

```{r}
# You can pull out just certain variables as well  
# This results in the same thing as above

flightdata %>% 
  select(-year)

```

Let's put all this together now, and use it to isolate our Sarasota flights.

```{r}

flightdata %>% 
  filter(dest == "SRQ", month <= 3) %>% 
  select(carrier, flight, origin, dest, 
         month, day, sched_dep_time, dep_delay)

```

Perfect.

Now remember of course, R is only doing what we're asking here: it's returning the slice of the data we've asked for...but it hasn't saved it anywhere.

So if our goal was to focus in on just these first-quarter Sarasota flights from New York, we might want to save the results in a new dataframe object. Let's do that.

```{r}

sarasota_q1 <- flightdata %>% 
  filter(dest == "SRQ", month <= 3) %>% 
  select(carrier, flight, origin, dest, 
         month, day, sched_dep_time, dep_delay)

```

Did it work? It did. Look at the upper right environment box.

Why don't we actually see anything show up below? We didn't ask R to actually show us anything that we saved. We can do that by, again, simply using the name of our new dataframe.

```{r}

sarasota_q1

```

There we go.

## Saving data to a file

Now a key point to remember here is that we've now saved our Sarasota flight data to our R session. And any time we come back later and run this set of steps in the script, we can isolate Sarasota in a similar way.

But -- this doesn't mean we've actually saved the data to a *file*. Which is cool, as sometimes (many times really) we might not need or want to do that.

However, if we do...let's say to use this data later in another script, or to share what we found with others, we need to get the new dataset out of R and into a file saved to our computer itself.

So how do we do *that*?

```{r}
#special R binary format
saveRDS(sarasota_q1, "data/sarasota_q1_flights.rds")

```

We can also save to a whole host of other file formats, some of which you may have heard of, and other not. Let's pick one that you have: Excel.

Saving to Excel is as simple as the following using the `writexl` package...

```{r}

write_xlsx(sarasota_q1, "data/sarasota_q1_flights.xlsx")

```

Now that we've saved our New York to Sarasota flights as a file, let's use that file for the next lesson.
