---
title: "Lab 01 - Grouping/Filtering"
author: Paulina Lopez, Sophie Holtzman, Sarina Oshiro
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).
When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the slice of WinRed contribution data from West Virginia.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
wv_winred_contribs <- read_rds("data/lab_01.rds")
```

## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.
**Q1** How many rows and columns does this dataframe have?
**A1** Rows: 13,382; Columns: 24

```{r}
glimpse(wv_winred_contribs)
```

## Answer questions

**Q2** Which West Virginia city provided the largest number of contributions? How many?
**A2** Martinsburg, with 781 contributions.

```{r}

wv_winred_contribs %>%
  group_by(city) %>%
  summarise(count_contribs = n()) %>% 
  arrange(desc(count_contribs)) %>% 
  View()



```
**Q3** Which West Virginia city accounted for the fewest contributions? How many?
**A3** There are 40 cities with tied for fewest contributions (Auburn-->Unger)

```{r}

wv_winred_contribs %>%
  group_by(city) %>%
  summarise(count_contribs = n()) %>% 
  arrange(count_contribs) %>% 
  View()


```
**Q4** What's your best educated guess to explain why the city identified in question 2 had more contributions compared to the cities identified in question 3? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.  
**A4** Martinsburg is the 6th largest city in West Virginia, so it makes sense that the most contributions via WinRed came from this city. The other top 5 major cities are more Democratic, so they would not have made as many contributions to Republicans. 
Source: https://worldpopulationreview.com/states/cities/west-virginia


**Q5** What recipient got the largest total amount of money? How many contributions did they receive and what was the average donation?
**A5** SAVE AMERICA JOINT FUNDRAISING COMMITTEE received the largest amount of total money, at 66972.89. They recieved 2,448 contributions. The average was contribution was 27.36.

```{r}

wv_winred_contribs %>%
  group_by(memo_text) %>%
  summarise(
    count_contribs = n(),
    max_amount = max(amount),
    mean_amount = mean(amount),
    total_amount = sum(amount)
  ) %>%
  arrange(desc(total_amount))




```

**Q6** What is that top recipient and what explains its popularity? Hint: this will require web research, you won't be able to answer this purely from the data set.  You do not need to write any code for this question, so no codeblock is provided.  
**A6** The Save America Joint Fundraising Committee was the fundraising arm for Donald Trump. He is a large national candidate, so this makes sense that the most money was contributed to that account. 


**Q7**. Produce a summary table that has the following information for West Virginia as a whole (the whole state, not by city):
* count of contributions
* total amount of all contributions
* the average contribution amount
* the median contribution amount
* the smallest contribution amount
* the largest contribution amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings.

**A7**. 
The total amount of contributions for the entire state of WV jumped out at us as a newsworthy nugget. This table also produced data on the minimum and maximum contributions, which we found interesting because someone donated only 1 cent.
```{r}

wv_summary <- wv_winred_contribs %>%
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount), 
    min_amount = min(amount),
    max_amount = max(amount)
  ) %>%
  arrange(desc(total_amount)) %>% 
  View()


```
